---
title: "Walkthrough reading in synthetic OMOP data for Critical Care"
output:
  html_document:
    keep_md: true
author: Andy South
date: "2024-12-03"
---

<!-- 
  *.md is generated from `*.Rmd` in /dynamic-docs/, this is updated in a github action,
  you can regenerate this by running the "Find and knit Rmd files" step from the github action.
-->


This document is an introductory walkthrough, demonstrating how to read into R some OMOP data from a series of csv files, join on concept names and explore and visualise the data.

It uses synthetic data created in UCLH as a part of the [Critical Care Health Informatics Collaborative (CCHIC) project](https://safehr-data.github.io/uclh-research-discovery/projects/uclh_cchic_s0/index.html).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Installing & loading required packages

If any of the `library($PACKAGE)` packages aren't installed, you can install them by running `install.packages("$PACKAGE")`,
replacing `$PACKAGE` with the package name.

```{r packages}

# install omopcept from Github if not installed
if (!requireNamespace("omopcept", quietly=TRUE)) 
{
  if (!requireNamespace("remotes", quietly=TRUE)) install.packages("remotes")
  
  remotes::install_github("SAFEHR-data/omopcept")
}

library(readr)
library(dplyr)
library(here)
library(gh)
library(omopcept)
library(ggplot2)
library(stringr)
library(lubridate)

```

## Downloading & reading in OMOP data 

Here we will download & read in some UCLH critical care data stored in Github (if they are not already present from a previous download).

```{r download-omop}

repo <- "SAFEHR-data/uclh-research-discovery"
path <- "_projects/uclh_cchic_s0/data"
destdata <- here("dynamic-docs/02-omop-walkthrough-critical-care/data")

# only download if not already present
if (! file.exists(file.path(destdata,"person.csv")))
{
  # Make GitHub API request to list contents of given path
  response <- gh::gh(glue::glue("/repos/{repo}/contents/{path}"))

  # Download all files to the destination dir
  purrr::walk(response, ~ download.file(.x$download_url, destfile = file.path(destdata, .x$name)))

  list.files(destdata)
}


```
## Reading in the OMOP data 

We now have the OMOP data as a series of csv files in a folder.

The [omopcept package](https://github.com/SAFEHR-data/omopcept) developed at UCLH, and installed above, has a function for reading in OMOP tables to a single list object.

You could alternatively use [CDMConnector::cdm_from_files](https://darwin-eu.github.io/CDMConnector/reference/cdm_from_files.html) to access OMOP data from files like this and [CDMConnector::cdm_from_con](https://darwin-eu.github.io/CDMConnector/reference/cdm_from_con.html) if your OMOP data are in a database. We aim to add documentation about that here soon.

```{r read-omop}

omop <- omopcept::omop_cdm_read(destdata, filetype="csv")

# names() can show us names of the tables read in
names(omop)

```

## Looking at the `person` table 

The OMOP tables are stored as data frames within the list object & can be accessed by the table name.

Thus we can find the names of the columns in `person`, use `glimpse()` to preview the data and `ggplot()` plot some of them. Note that not all columns contain data and in that case are filled with `NA`.

```{r explore-person}

# names() can also show column names for one of the tables
names(omop$person)

# glimpse table data
glimpse(omop$person)

# plot some columns, patient birth years by gender
ggplot(omop$person, aes(x=year_of_birth, fill = as.factor(gender_concept_id))) +
  geom_bar() +
  theme_minimal()

```

In the plot above bars are coloured by `gender_concept_id` which is the OMOP ID for gender, but we don't actually know which is which. We will look at resolving that by retrieving OMOP concept names in the next section.

## Getting names for OMOP concept IDs

To get the names for these and other concept IDs we need to use the OMOP vocabularies that store concept IDs and names. In some cases an OMOP database will contain the vocabularies as extra tables. In this case the vocabularies are not provided with the csv files.

One way to add concept names is to use a function called `omop_join_name_all()` from the [omopcept package](https://github.com/SAFEHR-data/omopcept). It will add columns containing concept names for all columns identified as containing concept ids (based on the column name). It works on single tables but will also accept a list of tables and add name columns to all of them (it can take a good few seconds to join on all the name columns).

```{r omop-join-names-all}

# join name columns onto all tables
omop_named <- omop |> omop_join_name_all()

# the names columns that have been added
names(omop_named$person) |> str_subset("name")

# now the gender name column can be used in the plot
ggplot(omop_named$person, aes(x=year_of_birth, fill=as.factor(gender_concept_name))) +
  geom_bar() +
  theme_minimal()

```

## Looking at the `measurement` table

We can use the `measurement_concept_name` column (that was added by `omop_join_name_all()` above) to see which are the most common measurements.

```{r explore-measurement}

glimpse(omop_named$measurement)

# most frequent measurement concepts
count(omop_named$measurement, measurement_concept_name, sort=TRUE)

```

## Looking at the `observation` table

We can use the `observation_concept_name` column (that was added by `omop_join_name_all()` above) to see which are the most common observations.

```{r explore-observation}

glimpse(omop_named$observation)

# most frequent observation concepts
count(omop_named$observation, observation_concept_name, sort=TRUE)

```

## Looking at the `drug_exposure` table

We can use the `drug_concept_name` column (that was added by `omop_join_name_all()` above) to see which are the most common drugs.

```{r explore-drug_exposure}

glimpse(omop_named$drug_exposure)

# most frequent drug_exposure concepts
count(omop_named$drug_exposure, drug_concept_name, sort=TRUE)

```

## Looking at the `visit_occurrence` table

The `visit_occurrence` table contains times and attributes of visits. Other tables (e.g. `measurement` & `observation`) have a `visit_occurrence_id` column that can be used to establish the visit that they were associated with. Visits have a start & end date, in these synthetic data the interval between them can be substantial.

```{r explore-visit_occurrence}

glimpse(omop_named$visit_occurrence)

# plot timeline of visit starts
omop_named$visit_occurrence |>
  #ggplot(aes(x=visit_start_date)) + geom_bar() +
  #replacing above with year tally otherwise bars too narrow to be reliably visible
  group_by(visit_concept_name, year = lubridate::floor_date(visit_start_date, "year")) |>
  summarize(nvisits = n()) |>
  ungroup() |> 
  ggplot(aes(x = year, y = nvisits, fill=visit_concept_name)) +
  geom_col() +
  facet_grid(vars(as.factor(visit_concept_name))) +
  theme_minimal() +
  theme(legend.position = "none")

```


## Joining `person` data to other tables

The OMOP common data model is person centred. Most tables have a `person_id` column that can be used to relate these data to other attributes of the patient. Here we show how we can join the `measurement` and `person` tables to see if there is any gender difference in measurements. A similar approach could be used to join to other tables including `observation` & `drug_exposure`.

```{r join-person-measurement}

joined_mp <- omop_named$measurement |> 
  left_join(omop_named$person, by="person_id")

freq_top_measures <- joined_mp |> 
  count(measurement_concept_name,gender_concept_name, sort=TRUE) |> 
  filter(n > 1) |> 
  #TODO not sure why collect() needed here
  #but without it the following ggplot erros with
  #Error in `fortify()`:! `data` must be a <data.frame>, or an object coercible by
  #something to do with arrow tables ?
  collect()

#plot
freq_top_measures |> 
  ggplot(aes(y=measurement_concept_name, x=n, fill=as.factor(gender_concept_name))) +
    geom_col() +
    facet_wrap(vars(as.factor(gender_concept_name))) +
    theme_minimal() +
    theme(legend.position = "none")
    

```

Note that we use `left_join` here because we only want to join on `person` information for rows occurring in the `measurement` table which is the left hand argument of the join. Also note that in this example we end up with one row per patient because the synthetic `measurement` table only has one row per patient. Usually we would expect multiple measurements per patient that would result in multiple rows per patient in the joined table.

## Differences between these synthetic data and real patient data

These particular synthetic data are useful to demonstrate the reading in and manipulation of OMOP data but there are some major differences between them and real patient data.

1. `person`, `measurement`, `observation` & `drug_exposure` tables are all same length (100 rows), in real data one would expect many more measurements, observations & drug exposures than patients
2. Related to 1, in these data there are a single `measurement`, `observation` and `drug_exposure` per patient. In reality one would expect many tens or hundreds of these other values per patient.
